---
title: "HW6"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Study Question 2

A DAG encodes conditional independence between variables, and such relatinoship is implied by d-spearation.

\[ (Y \perp\!\!\!\perp_d X | S)_\text{G} \implies Y \perp\!\!\!\perp X | S \]

A set of variables $S$ d-separates $X$ and $Y$ in a DAG G when the variables in $S$ block all the paths between $X$ and $Y$. A path is blocked if and only if:

1. A non-collider in the subset of nodes is conditioned on, or
2. A collider in the subset of nodes is not contitioned on and has no descendents which is conditioned on.

The 'dagitty' package applies these criteria for d-separation to identify the testable implications of conditional indenpendence between two variables in a DAG.

### Part A

Here we plot the DAG of Peal's example and print all the testable conditional independence relationships implied by this DAG.

```{r part a}
library(dagitty)
library(ggdag)

G1 = dagitty(
  'dag{
    Z1 [pos = "-1, -1"]
    X1 [pos = "0, -1"]
    Z2 [pos = "-1, 1"]
    X3 [pos = "0, 1"]
    Y [outcome, pos = "1, 1"]
    D [exposure, pos = "1, -1"]
    M [mediator, pos = "1, 0"]
    X2 [pos = "0, 0"]
    Z1 -> X1
    X1 -> D
    Z1 -> X2
    Z2 -> X3
    X3 -> Y
    Z2 -> X2
    D -> Y
    X2 -> Y
    X2 -> D
    M -> Y
    D -> M
  }'
)
ggdag(G1) + theme_dag()
print(impliedConditionalIndependencies(G1))
```

### Part B

Please see below the DAG with variables $D$, $Y$, $X_2$ and $X_3$ as well as testable restrictions implied by this DAG.

```{r part b}
G2 = dagitty(
  'dag{
    X3 [pos = "0, 1"]
    Y [outcome, pos = "1, 1"]
    D [exposure, pos = "1, 0"]
    X2 [pos = "0, 0"]
    X3 -> Y
    D -> Y
    X2 -> Y
    X2 -> D
  }'
)
ggdag(G2) + theme_dag()
print(impliedConditionalIndependencies(G2))
```

### Part C

Please see below the DAG with variables $D$, $Y$ and $X_2$. There is no testable restriction implied by this DAG.

```{r part c}
G3 = dagitty(
  'dag{
    Y [outcome, pos = "1, 1"]
    D [exposure, pos = "1, 0"]
    X2 [pos = "0, 0"]
    D -> Y
    X2 -> Y
    X2 -> D
  }'
)
ggdag(G3) + theme_dag()
print(impliedConditionalIndependencies(G3))
```

### Part D

Please see below the DAG with all the variables in Pearl's example except for $X_3$ as well as testable restrictions implied by this DAG.

```{r part d}
G4 = dagitty(
  'dag{
    Z1 [pos = "-1, -1"]
    X1 [pos = "0, -1"]
    Z2 [pos = "-1, 1"]
    Y [outcome, pos = "1, 1"]
    D [exposure, pos = "1, -1"]
    M [mediator, pos = "1, 0"]
    X2 [pos = "0, 0"]
    Z1 -> X1
    X1 -> D
    Z1 -> X2
    Z2 -> X2
    D -> Y
    X2 -> Y
    X2 -> D
    M -> Y
    D -> M
  }'
)
ggdag(G4) + theme_dag()
print(impliedConditionalIndependencies(G4))
```

### Part E

Please see below the DAG with the arrow between $D$ and $X_2$ reversed as well as testable restrictions implied by this DAG.

```{r part e}
G5 = dagitty(
  'dag{
    Z1 [pos = "-1, -1"]
    X1 [pos = "0, -1"]
    Z2 [pos = "-1, 1"]
    X3 [pos = "0, 1"]
    Y [outcome, pos = "1, 1"]
    D [exposure, pos = "1, -1"]
    M [mediator, pos = "1, 0"]
    X2 [pos = "0, 0"]
    Z1 -> X1
    X1 -> D
    Z1 -> X2
    Z2 -> X3
    X3 -> Y
    Z2 -> X2
    D -> Y
    X2 -> Y
    D -> X2
    M -> Y
    D -> M
  }'
)
ggdag(G5) + theme_dag()
print(impliedConditionalIndependencies(G5))
```

We may undertake the Verma-Pearl test for linear ASEMs to distinguish between Pearl's example and the modified DAG. In Pearl's example, $X_3$ is d-separated from $D$ when $X_2$ is conditioned on. However, in the modified DAG, $X_3$ is d-separated from $D$ when $X_2$ is not conditioned on. So, consider the hypothesis for linear ASEMs that $X_3$ is independent of $D$ given $X_2$. If the coefficient $\alpha = 0$ in the following equation,
\[ X_3 = \alpha D + \beta X_2 + \epsilon \quad (\epsilon \perp X_2)  \]
Pearl's example should be an appropriate model as opposed to the modified DAG.